\textbf{Differences in COVID-19 vaccination in the province of Ontario across Health Regions and socio-economic strata}

### 2023-07-07 manuscript_v6/01-background.qmd
updated discussion and conclusion, acknowlegements

Reasons given for this inequality in vaccination uptake have included medical mistrust due to systemic medical racism [@stoler2021;@mosby2021], mistrust in vaccines [@willis2021], and the existence of health misinformation and disinformation [@bogart2021;@mosby2021;@freeman2020]. However, it is important also to consider that vaccination uptake can be influenced by geographical (spatial) factors. In this regard, differences in COVID-19 vaccination rates have been associated with varied regional attitudes towards vaccination [@malik2020], lack of vaccine access in areas where underrepresented groups reside[@bogoch2022], as well as an absence of prioritization of areas inhabited by vulnerable groups [@nguyen2021]. Other studies have also shown heterogeneity in vaccine uptake within small governmental administrative units such as counties [@mollalo2021; @yang2022; @tiu2022; @bhuiyan2022], indicating that accounting for geographical differences in vaccination can help predict patterns of booster uptake [@wood2022]. However, there is a limited amount of studies that have analyzed geographical differences in vaccination in Canada. Existing studies in this area have focused on differences within certain cities, such as Toronto [@choi2021], or Montréal [@mckinnon2021]), or have explored differences between provinces [@guay2022], but to our knowledge, there are no studies that analyze  differences in vaccination uptake at the provincial level. Such need is specially important in the context of Canada's pandemic response goals, which have been to minimize serious illness and deaths while minimizing societal disruption [@naci2023]. Analyzing differences in vaccination uptake within the provinces can aid to identify inequalities that might exist and that need to be addressed before the advent of another pandemic.

>>>>

Reasons given for this inequality in vaccination uptake have included medical mistrust due to systemic medical racism [@stoler2021;@mosby2021], mistrust in vaccines [@willis2021], and the existence of health misinformation and disinformation [@bogart2021;@mosby2021;@freeman2020]. However, it is important also to consider that vaccination uptake can be influenced by geographical (spatial) factors. In this regard, differences in COVID-19 vaccination rates have been associated with varied regional attitudes towards vaccination [@malik2020], lack of vaccine access in areas where underrepresented groups reside[@bogoch2022], as well as an absence of prioritization of areas inhabited by vulnerable groups [@nguyen2021]. Other studies have also shown heterogeneity in vaccine uptake within small governmental administrative units such as counties [@mollalo2021; @yang2022; @tiu2022; @bhuiyan2022], indicating that accounting for geographical differences in vaccination can help predict patterns of booster uptake [@wood2022]. However, there is a limited amount of studies that have analyzed geographical differences in vaccination in Canada. Existing studies in this area have focused on differences within certain cities, such as Toronto [@choi2021], or Montréal [@mckinnon2021]), or have explored differences between provinces [@guay2022], but to our knowledge, there are no studies that analyze differences in vaccination uptake at the intra-provincial level. Such need is specially important in the context of Canada's pandemic response goals, which have been to minimize serious illness and deaths while minimizing societal disruption [@naci2023]. Analyzing differences in vaccination uptake within the provinces can aid to identify inequalities that might exist and that need to be addressed before the advent of another pandemic.

### 2023-07-07 manuscript_v6/04-discussion.qmd
updated discussion and conclusion, acknowlegements

In this study, we hypothesized that differences in COVID-19 vaccination uptake were present between the Health Regions between late 2021 and early 2022. Our goal was to determine which socio-demographic groups could be impacted by these disparities in order to provide decision-makers with information that could be used to develop policies focused on reducing or eliminating these differences and ensuring that the Health Region model is able to fulfill its mission of improving health access for all Ontarians.

Our results show that indeed, there were differences in vaccination odds across Ontario in certain socio-demographic groups. Specifically, those who identified as Arab/Middle Eastern, Black, Latin American, having mixed racial or ethnic background, or that belonged to other groups not explicitly included in the survey (Southeast Asian, Filipino, West Asian, and minority groups not identified elsewhere) had vaccination odds that were between a third and a half of that of individuals that identified as White or Caucasian (@fig-models). These results are consistent with previous studies that have shown lower vaccination rates in individuals with the same socio-demographic characteristics [@guay2022;@muhajarine2021;@carter2022;@hussain2022].

@@

This result is likely reflects in part the fact that individuals in underrepresented groups tend to perform occupations that have been deemed as "essential" in the context of the pandemic [@hawkins2020;@ct2021], which include workers in the areas of grocery stores, gas stations, warehouses, distribution, and manufacturing, all being occupations for which an income within the significant brackets identified in the analysis is to be expected. In Ontario, these workers had priority for COVID-19 vaccination [@mishra2021]; and there is evidence of interventions by vaccination staff in certain parts of the province to encourage vaccination uptake by these individuals [@gill2022]. These facts, combined with evidence of increased trends in vaccination in this group elsewhere [@nguyen2021b], suggest that the type of occupation of individuals in underrepresented groups played an important role in increasing the odds of vaccination.

@@

There are some limitations to the present study. First, the data collection design, which allowed respondents to withdraw from the survey at any point, and that deployed the questions in a random manner resulted in an elevated number of missing observations without a definite pattern and complicated the implementation of sensitivity analyses. Therefore, we focused on entries that had complete answers, and corrected the data using population-wide information from the Census. More granular corrections would be needed to obtain more accurate estimates. For example, our analysis identified higher odds of vaccination in the Central and West Health Regions, but in this case these differences are likely to be driven by the proportion of White/Caucasian individuals, who had higher vaccination rates than other racial groups. Correcting for each racial/ethnic group in each Health Region can provide a more accurate estimation of region-wide vaccination rates but unfortunately, at the moment this correction cannot be implemented as such stratification is has not been implemented in the Census.

>>>>

The COVID-19 pandemic has presented many significant challenges to the healthcare systems of the Canadian provinces [@alami2021]. In the case of Ontario, the recent adoption of the Health Region model at the onset of the pandemic signified an additional level of complexity because although PHO (through the 34 Public Health Units of the province) was the lead in vaccine distribution, in reality the vaccine rollout required active collaboration between PHO and OH in multiple instances [@ashcroft2023].

In this study, we hypothesized that there were differences in the uptake of the first dose of the COVID-19 vaccine between individuals living in the different Health Regions of Ontario during late 2021 and early 2022. Using the Health Regions as the base of our analysis was advantageous, as these new Health Regions match closely the geographical boundaries of the Health Regions that have been historically used to group Public Health Units in the province [@bolotin2018;@ohesi], thus providing information with relevance at both the public health and healthcare system level. Therefore, this analysis could provide an overall assessment of intra-provincial disparities that might need to be addressed moving forward by decision-makers, in order to ensure that in the future, and specially in the event of a future public health emergency, OH and PHO will able to collaboratively work in an efficient way and thus ensure that the population of the province benefits in the long term of the newly implemented Health Region model.

<!-- Specifically, our goal was to determine which socio-demographic groups could be impacted by these differences in vaccination rates in order to highlight challenges that will need to be addressed by OH to determine the exact nature of its future interactions with PHO in the event of a future public health emergency. -->

<!-- The complex relationships within the health system of Ontario, with the recent implementation of OH make necessary to determine  COVID-19 pandemic highlighted the need of an active collaboration between PHO and OH -->

<!-- The recent implementation of the Health Region model  vaccination campaign in Ontario  -->

<!-- Due to the recent implementation of the Health Region model in the province, and the Our goal was to determine which socio-demographic groups could be impacted by these disparities, in order to provide decision-makers with information that could be used to inform  -->

<!-- develop policies focused on reducing or eliminating these differences and ensuring that the Health Region model is able to fulfill its mission of improving health access for all Ontarians. -->

<!-- In this study, we hypothesized that differences in the uptake of the first dose of the COVID-19 vaccine were present between the Health Regions between late 2021 and early 2022. Our goal was to determine which socio-demographic groups could be impacted by these disparities in order to provide decision-makers with information that could be used to develop policies focused on reducing or eliminating these differences and ensuring that the Health Region model is able to fulfill its mission of improving health access for all Ontarians. -->

Our results show that indeed, there were differences in the uptake of the first dose of the vaccine across Ontario in certain socio-demographic groups. Specifically, those who identified as Arab/Middle Eastern, Black, Latin American, having mixed racial or ethnic background, or that belonged to other groups not explicitly included in the survey (Southeast Asian, Filipino, West Asian, and minority groups not identified elsewhere) had vaccination odds that were between a third and a half of that of individuals that identified as White or Caucasian (@fig-models). These results are consistent with previous studies that have shown lower vaccination rates in individuals with the same socio-demographic characteristics [@guay2022;@muhajarine2021;@carter2022;@hussain2022].

@@

This result likely reflects in part the fact that individuals in underrepresented groups tend to perform occupations that have been deemed as "essential" in the context of the pandemic [@hawkins2020;@ct2021], which include workers in the areas of grocery stores, gas stations, warehouses, distribution, and manufacturing, all being occupations for which an income within the significant brackets identified in the analysis is to be expected. From one side, individuals in essential occupations in the province experienced higher rates of morbidity and mortality during the first year of the pandemic [@rao2021], but later on, they had priority for COVID-19 vaccination [@mishra2021]. Additionally, it is known that vaccination uptake in these individuals was encouraged by vaccination staff in certain parts of the province [@gill2022]. These facts, combined with evidence of increased trends in vaccination in this group elsewhere [@nguyen2021b], suggest that in Ontario, the type of occupation of individuals in underrepresented groups (which might have also affected their decision to get a vaccine based on their knowledge of increased risk), played an important role in the higher the odds of vaccination observed in these individuals.

@@

There are some limitations to the present study. First, the data collection design, which allowed respondents to withdraw from the survey at any point, and that deployed the questions in a random manner resulted in an elevated number of missing observations without a definite pattern and complicated the implementation of sensitivity analyses. Therefore, we focused on entries that had complete answers, and corrected the data using population-wide information from the Census. However, more granular corrections would be needed to obtain more accurate estimates: For example, our analysis identified higher odds of vaccination in the Central and West Health Regions, but in this case these differences are likely to be driven by the proportion of White/Caucasian individuals, who had higher vaccination rates than other racial groups. Correcting for each racial/ethnic group in each Health Region would provide a more accurate estimation of region-wide vaccination rates but unfortunately, presently this correction cannot be implemented as such stratification has not been implemented in the data that can be obtained from the Census.

### 2023-07-07 manuscript_v6/05-conclusion.qmd
updated discussion and conclusion, acknowlegements

The implementation of the Health Regions in Ontario aimed at reducing the bureaucratic complexity and health disparities identified under the LHIN model. However, there are currently multiple challenges that need to be addressed to ensure that the new model can improve healthcare for the inhabitants of the province. First, the fact that each Health Region now covers a large geographical area that was served by multiple LHINs in the past creates a complex socio-demographic landscape that is different in each case due the different levels of rurality and representation of equity-deserving groups that are now within each Health Region. So far, the evidence collected during the COVID-19 pandemic indicates that differences in vaccination uptake are associated to a lack of infrastructure and resources that can adequately support personalized care to marginalized individuals. In the near future, health decision-makers will need to consider the implementation of policies that are focused on addressing this problematic.

Moreover, the recent nature in the adaption of the Health Region poses a challenge for researchers in the acquisition of data and information that can be used to analyze the performance of the new model. From one side, the Health Regions have not been incorporated as part of Census data (LHINs were considered before in the Census), and this impact the amount and level of detail of available information. On the other hand, evaluations of the Health Region model are at the moment not considered in the Annual Reports of the Auditor General of Ontario, which was critical in the past to identify limitations in the LHIN model. Currently, the only demographic information available for each Health Region is provided by Ontario Health (the agency that administers the Health Regions) but this information only provides general estimates that do not allow for detailed analyses of performance indicators (such as hospitalizations, readmissions, and trends in chronic disease incidence) between the Regions. Therefore, there is a pressing need for open data that can be used by researchers and decision-makers to examine the performance of the Health Region model.

>>>>

The implementation of the Health Regions in Ontario aimed at reducing the bureaucratic complexity and health disparities identified under the LHIN model. However, there are certain challenges that this new model will need to overcome in order to work syngergistically with the rest of the healthcare system of the province. First, each Health Region now covers a large geographical area that was served in the past by multiple LHINs. This in turn creates a complex socio-demographic landscape within each Health Region that is likely to be different in each case due the different levels of rurality and the proportion of underrepresented groups in each Health Region.

Second, it is likely that a future pandemic or public health emergency will impact the province in the years ahead. Although broadly speaking, PHO and OH have been able to collaborate on the COVID-19 vaccination rollout, future events will require a unified response from both the public health arm of the government (represented by PHO), and the healthcare system agency (OH). So far, the COVID-19 pandemic has as a stress test for both agencies, showing the necessity from decision makers to clearly define how PHO and OH will interact in the future in order to ensure that in the future, Ontarians can benefit by virtue of a robust collaboration between both agencies. The evidence collected in this study shows that healthcare disparities experienced by underrepresented individuals in the province are the result of complex relationships between infrastructure, resources, occupations, geography, and socio-economic factors, among others. Therefore, this complexity presents an ongoing challenge for the healthcare system of Ontario, which will need to adapt and continue to aim for an integrative approach that can serve those that are most vulnerable.

<!-- So far, the evidence collected during the COVID-19 pandemic indicates that differences in vaccination uptake are associated to a lack of infrastructure and resources that can adequately support personalized care to marginalized individuals. In the near future, health decision-makers will need to consider the implementation of policies that are focused on addressing this problematic. -->

Finally, the recent nature in the adaption of the Health Region poses a challenge for researchers in the acquisition of data and information that can be used to analyze the performance of the new model. From one side, the Health Regions have not been incorporated as part of Census data (LHINs were considered before in the Census), and this impact the amount and level of detail of available information. On the other hand, evaluations of the Health Region model are at the moment not considered in the Annual Reports of the Auditor General of Ontario, which was critical in the past to identify limitations in the LHIN model. Currently, the only demographic information available for each Health Region is provided by Ontario Health (the agency that administers the Health Regions) but this information only provides general estimates that do not allow for detailed analyses of performance indicators (such as hospitalizations, readmissions, and trends in chronic disease incidence) between the Regions. Therefore, there is a pressing need for open data that can be used by researchers and decision-makers to examine the performance of the Health Region model.
